SparseIDS: Learning Packet Sampling with Reinforcement Learning

### 2020-02-14 paper_rl/paper.tex
Paper fixes

\glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data. They allow determining if a flow is malicious or not already before it is over, making it possible to take action immediately. However, considering the large number of packets that have to be inspected, the question of computational efficiency arises. We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high. Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy by learning to sample only relevant packets. A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can even skip arbitrarily many packets in one step. This enables saving even more computational resources for long sequences. Inspecting \ours{}'s behavior of choosing packets shows that it adopts different sampling strategies for different attack types and network flows. Finally we build an automatic steering mechanism that can guide \ours{} in deployment to achieve a desired level of sparsity.

@@

The classification reward metric (\autoref{eq:R_classification}) captures how correct the classifier is on average for all future packets of a flow. The label of each packet in a flow has a value of 0 for non-attack traffic and 1 for attack traffic with one flow being only attack or non-attack but not mixed. The confidence is the sigmoided $\left(\frac{e^x}{1+e^x}\right)$ output of the classifier, yielding a value between 0 and 1, where 0 is absolute confidence for a flow being benign and 1 is absolute confidence for a flow being an attack.
\begin{align}
\begin{split}
R_{\text{classification},n} = \frac{1}{N-n} \sum_{i=n+1}^{N} |1 - \left(\text{label}_i - \text{confidence}_i\right)|
\end{split}
\label{eq:R_classification}
\end{align}

Intuitively, the classification reward makes sure that packets are chosen that contain the most information and thus result in the highest classification performance.

The sparsity reward metric (\autoref{eq:R_sparsity}) captures how much sparsity is achieved on average for all future packets of a flow. E.g. for a flow of length 10 if after 5 packets have been processed, 3 packets are chosen and 2 are skipped from the remaining 5 packets, the sparsity reward metric at packet 5 is $\frac{2}{10-(4+1)}=\frac{2}{5}$.
%\note{blue}{Fares: This equation can be shortened by diving by the total count of packets left.\\
%Fares: If the length of the flow is not known, how do we estimate T}.
%\note{red}{Max: It's only relevant during training when you know the lengths of the flows. In deployment, you don't need the rewards anymore...}
\begin{align}
\begin{split}
R_{\text{sparsity},n} = \frac{\sum_{i=n+1}^{N} 1_{\text{skipped},i}}{N-(n+1)}
\end{split}
\label{eq:R_sparsity}
\end{align}

\autoref{fig:rewards} shows how classification and sparsity rewards are computed for a toy example.

@@

Combining these objectives, the actor network aims to minimize the loss function
\begin{align}
\begin{split}
l_{\text{a},n} =& \underbrace{ -\log \left( \pi \left( a_n \given s_n ; \theta_\text{a} \right) \right) U_n}_{\text{Policy Loss}} \\
&- \beta \underbrace{H\left( \pi\left( s_n; \theta_a \right)\right) }_{\text{Entropy}}
\end{split}\label{eq:actor}
\end{align}

@@

\begin{figure}[h]
\centering
  \includegraphics[width=0.8\columnwidth]{img/rnn-sampling.pdf}
  \caption{SparseIDS framework. Three deep networks are shown: the LSTM classifier, the actor network and the critic network. The actor takes actions based on the feedback from the the critic network and the classifier. The critic network tracks the actor's actions based on the classifier confidence and finally, the packets selector parses packets based on the decision of the actor network.}
  \label{fig:sampling}
\end{figure}

@@

The architecture in \autoref{fig:neuralNetworkArchitecture} shows that the actor network has two outputs: $\mu$ and $\sigma$. These encode the mean and standard deviation of a log-normal distribution and can never be smaller or equal to zero and hence we always apply a rectifier function at the output. We choose the softplus, which is a smooth rectifier function:
\begin{align}
\text{softplus}(x) = \log\left(1 + e^x \right)
\end{align}

When sampling from the log-normal distribution, a real number is obtained. We round it down and then add 1. This gives us a natural number between 1 and $\infty$, which states how many packets should be skipped.

When experimenting with continuous actions, we encountered one practical problem: At the beginning of training, the classifier would still perform badly and would not give the actor a useful reward. As a result, the actor would only focus on sparsity and aim to maximize it, as the classification reward would always be equally bad no matter how many packets are chosen. Thus, the actor would learn to let $\mu$ go to infinity. This would increase the reward at the beginning when the interaction between the three components of our \gls{rl} system is still fragile. As a solution, we add a penalty to the reward calculation when the actor jumps far behind the last flow packet. The modified reward applies only to the last chosen packet and penalizes the actor jumping more than one packet behind the last packet. The sparsity reward formula (\autoref{eq:R_sparsity}) of the last packet is modified as follows, where \textit{last} refers to the index of the last chosen packet in the flow ($n=\textit{last}$):

\begin{align}
\begin{split}
R_{\text{sparsity},\textit{last}} = \frac{\left(\sum_{i=\textit{last}+1}^{N} 1_{\text{skipped},i}\right)}{N- (\textit{last}+1+\left(\textit{last}+a_{\textit{last}} - N - 1\right))}
\end{split}
\label{eq:R_sparsity_last}
\end{align}

Specifically, it adds the term $\left(\textit{last}+a_{\textit{last}} - N - 1\right)$ to the denominator which penalizes the actor jumping more than one packet behind the flow. Jumping just right after the flow does not get penalized because if the actor considers the rest of the flow as irrelevant, ignoring the rest of the flow and jumping to the first packet behind the flow is a useful action.

@@

%\begin{table}[h]
%\caption{Occurrence frequency of the attack flows.}
%\label{tab:occurrence}
%\centering
%\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
%]{
%\begin{tabular}{l r}
%\toprule
%Attack type & \hspace*{-4mm}Proportion \\ \midrule
%DoS Hulk	&	10.10\%	\\
%PortScan, Firewall	&	6.90\%	\\
%DDoS LOIT	&	4.08\%	\\
%Infiltration	&	3.30\%	\\
%DoS GoldenEye	&	0.32\%	\\
%DoS SlowHTTPTest	&	0.18\%	\\
%DoS Slowloris	&	0.17\%	\\
%Brute-force SSH	&	0.11\%	\\
%Botnet ARES	&	0.03\%	\\
%XSS attack	&	0.03\%	\\
%PortScan, no Fw.	&	0.02\%	\\
%Brute-force FTP	&	0.01\%	\\
%SQL injection	&	$<$0.01\%	\\
%Heartbleed	&	$<$0.01\%	\\
%\bottomrule
%\end{tabular}
%}{}
%\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
%]{
%\begin{tabular}{l r}
%\toprule
%Attack type & \hspace*{-4mm}Proportion \\ \midrule
%Exploits	&	1.42\%	\\
%Fuzzers	&	1.01\%	\\
%Reconnaissance	&	0.58\%	\\
%Generic	&	0.21\%	\\
%DoS	&	0.19\%	\\
%Shellcode	&	0.08\%	\\
%Analysis	&	0.03\%	\\
%Backdoors	&	0.02\%	\\
%Worms	&	0.01\%	\\
%\bottomrule
%\end{tabular}
%}{}
%\end{table}

For our experiments, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset, which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks. The distribution of different attacks in the dataset are shown in \autoref{tab:occurrence}.

@@

%\begin{table}
%\caption{Comparing classification metrics per flow with different tradeoffs for both CIC-IDS-2017 and UNSW-NB15.} \label{tab:results_tradeoff}
%\centering
%\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
%& \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
%Tradeoff & 0.0 & 0.1 & 0.5 & 0.0 & 0.1 & 0.5 \\	\midrule
%\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% & 0\% & 83.5\% & 84.5\% \\ \midrule
%\setrow{\bfseries}Accuracy & 99.5\% & 99.4\% & 99.3\% & 98.9\% & 98.4\% & 98.5\% \\
%Precision & 99.4\% & 98.5\% & 98.7\% & 81.1\% & 69.5\% & 70.8\% \\
%Recall & 98.5\% & 99.0\% & 98.6\% & 84.0\% & 89.9\% & 89.6\% \\
%F1 & 99.0\% & 98.7\% & 98.6\% & 82.5\% & 78.4\% & 79.1\% \\
%Youden & 98.4\% & 98.5\% & 98.2\% & 83.3\% & 88.6\% & 88.4\% \\
%\bottomrule
%\end{tabular*}
%\end{table}

\begin{table}[h]
% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
\vspace{0.075in}
%\parbox{.45\linewidth}{
\caption{Comparing classification metrics per flow with different tradeoffs. The higher the tradeoff, the higher the sparsity but also the lower the accuracy.} \label{tab:results_tradeoff}
\centering
\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
Tradeoff & 0.0 & 0.1 & 0.5 & 1.0 \\	\midrule
Sparsity & 0\% & 76.3\% & 76.5\% & \textbf{76.8\%} \\ \midrule
Accuracy & \textbf{99.5\%} & 99.4\% & 99.3\% & 99.0\% \\
Precision & \textbf{99.4\%} & 98.5\% & 98.7\% & 97.9\% \\
Recall & 98.5\% & \textbf{99.0\%} & 98.6\% & 98.1\% \\
F1 & \textbf{99.0\%} & 98.7\% & 98.6\% & 98.0\% \\
Youden & 98.4\% & \textbf{98.5\%} & 98.2\% & 97.4\% \\
\bottomrule
\end{tabular*}
%}
\end{table}

\begin{table}[h]
\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff 0.1) and alternative strategies.}
\label{tab:results_others}
\centering
\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
& \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
Accuracy & \textbf{99.4\%} & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
Precision & \textbf{98.5\%} & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
Recall & \textbf{99.0\%} & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
F1 & \textbf{99.7\%} & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
Youden & \textbf{98.5\%} & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
\bottomrule
\end{tabular*}
%}
\end{table}

\subsection{Influence of the tradeoff}

\autoref{tab:results_tradeoff} shows the influence of the tradeoff parameter $\alpha$. For columns with a tradeoff of 0.0, we train the classifier regularly with all packets without using \gls{rl}. For tradeoffs 0.1, 0.5 and 1.0 we train \ours{} using \gls{rl} with continuous actions and keep the tradeoff constant for the whole training. Results show that the higher the tradeoff, the fewer packets are sampled and the lower the accuracy.

@@

%\begin{table}
%\caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions for CIC-IDS-2017 and UNSW-NB15. The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
%\centering
%\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
%& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
%Actions & cont. & disc. & cont. & disc. \\	\midrule
%\setrow{\bfseries}Sparsity & 76.3\% & 73.6\% & 83.5\% & 86.6\% \\ \midrule
%\setrow{\bfseries}Accuracy & 99.4\% & 99.2\% & 98.4\% & 98.4\% \\
%Precision & 98.5\% & 98.7\% & 69.5\% & 68.0\% \\
%Recall & 99.0\% & 98.3\% & 89.9\% & 94.6\% \\
%F1 & 99.7\% & 98.5\% & 78.4\% & 79.2\% \\
%Youden & 98.5\% & 97.8\% & 88.6\% & 93.2\% \\
%\bottomrule
%\end{tabular*}
%\end{table}

\begin{table}[h]
% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
\vspace{0.075in}
\caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions. The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
\centering
\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
Actions & continuous & discrete \\	\midrule
Sparsity & \textbf{76.3\%} & 73.6\% \\ \midrule
Accuracy & \textbf{99.4\%} & 99.2\% \\
Precision & 98.5\% & \textbf{98.7\%} \\
Recall & \textbf{99.0\%} & 98.3\% \\
F1 & \textbf{99.7\%} & 98.5\% \\
Youden & \textbf{98.5\%} & 97.8\% \\
\bottomrule
\end{tabular}
\end{table}

\autoref{tab:results_disc_cont} compares continuous and discrete (20 different) actions and shows that continuous actions result in a higher accuracy with fewer samples being selected. This leads to the conclusion that at least for our use case and for the dataset we consider, continuous actions seem to perform better. Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while the only possibility for continuous actions is to increase or decrease the mean of the log-normal distribution, which can lead to worse performance. On the other hand, with discrete actions, the maximum number of samples to skip is fixed, which we consider to be a major drawback. Furthermore, training time is potentially higher since, for example, for a discrete action space of 20 actions, each option is treated as being completely different and, while actually there is not a large difference between skipping 18 or 19 packets, there is a big difference between skipping 0 packets compared to skipping 19 packets. This means that for discrete actions, all actions are perceived as being completely independent, which is not true as the reward function considers sparsity, which increases gradually as more packets are skipped and thus similar actions result in similar rewards.

@@

%\begin{table}
%\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff $0.1$) and alternative strategies.} \label{tab:results_others}
%\centering
%\begin{tabular}{>{\rowmac}l >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
%& \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
%\setrow{\bfseries}Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
%\setrow{\bfseries}Accuracy & 99.4\% & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
%Precision & 98.5\% & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
%Recall & 99.0\% & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
%F1 & 99.7\% & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
%Youden & 98.5\% & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
%\bottomrule
%\end{tabular}
%\end{table}

Next we compared the \gls{rl} technique used by \ours{} with several common sampling strategies suggested in \autoref{sec:other_sampling_techniques}.
\autoref{tab:results_others} reveals that \ours{} achieves the highest accuracy with the same sparsity. Not surprisingly, random sampling performs the worst, as it is likely that a flow is ``unfortunate'' and doesn't get enough packets sampled while another one gets more packets or the packets that carry more useful information. The best strategy besides the \gls{rl} approach seems to be ``first $m$'' sampling, which matches the conclusion from \cite{hartl_explainability_2020} that the first couple of packets are the most important ones for determining maliciousness of a flow.
``first $m$'' performs better than ``relative first $m$'' because for ``relative first $m$'', longer flows will have more packets sampled from them. Conversely, very short flows will get few packets. For example, with $p=0.2$, a flow of length 20 will get 4 packets while one with 4 packets will get only 1. However, with `first $m$'', the short and the long flow will get the same number of packets. Apparently it is better if all flows get the same number of packets than if long flows get more packets.

>>>>

\glspl{rnn} have been shown to be valuable for constructing \glspl{ids} for network data. They allow determining if a flow is malicious or not already before it is over, making it possible to take action immediately. However, considering the large number of packets that has to be inspected, the question of computational efficiency arises. We show that by using a novel \gls{rl}-based approach called \textit{\ours{}}, we can reduce the number of consumed packets by more than three fourths while keeping classification accuracy high. Comparing to various other sampling techniques, \ours{} consistently achieves higher classification accuracy by learning to sample only relevant packets. A major novelty of our \gls{rl}-based approach is that it can not only skip up to a predefined maximum number of samples like other approaches proposed in the domain of Natural Language Processing but can even skip arbitrarily many packets in one step. This enables saving even more computational resources for long sequences. Inspecting \ours{}'s behavior of choosing packets shows that it adopts different sampling strategies for different attack types and network flows. Finally we build an automatic steering mechanism that can guide \ours{} in deployment to achieve a desired level of sparsity.

@@

The classification reward metric (\autoref{eq:R_classification}) captures how correct the classifier is on average for all future packets of a flow. The label of each packet in a flow has a value of 0 for non-attack traffic and 1 for attack traffic with one flow being only attack or non-attack but not mixed. The confidence is the sigmoided $\left(\frac{e^x}{1+e^x}\right)$ output of the classifier, yielding a value between 0 and 1, where 0 is absolute confidence for a flow being benign and 1 is absolute confidence for a flow being an attack.
\begin{align}
\begin{split}
R_{\text{classification},n} = \frac{1}{N-n} \sum_{i=n+1}^{N} |1 - \left(\text{label}_i - \text{confidence}_i\right)|
\end{split}
\label{eq:R_classification}
\end{align}
Intuitively, the classification reward makes sure that packets are chosen that contain the most information and thus result in the highest classification performance.

The sparsity reward metric (\autoref{eq:R_sparsity}) captures how much sparsity is achieved on average for all future packets of a flow. E.g. for a flow of length 10 if after 5 packets have been processed, 3 packets are chosen and 2 are skipped from the remaining 5 packets, the sparsity reward metric at packet 5 is $\frac{2}{10-(4+1)}=\frac{2}{5}$.
%\note{blue}{Fares: This equation can be shortened by diving by the total count of packets left.\\
%Fares: If the length of the flow is not known, how do we estimate T}.
%\note{red}{Max: It's only relevant during training when you know the lengths of the flows. In deployment, you don't need the rewards anymore...}
\begin{align}
\begin{split}
R_{\text{sparsity},n} = \frac{\sum_{i=n+1}^{N} 1_{\text{skipped},i}}{N-(n+1)}
\end{split}
\label{eq:R_sparsity}
\end{align}
\autoref{fig:rewards} shows how classification and sparsity rewards are computed for a toy example.

@@

Combining these objectives, the actor network aims to minimize the loss function
\begin{align}
\begin{split}
l_{\text{a},n} =& \underbrace{ -\log \left( \pi \left( a_n \given s_n ; \theta_\text{a} \right) \right) U_n}_{\text{Policy Loss}} - \beta \underbrace{H\left( \pi\left( s_n; \theta_a \right)\right) }_{\text{Entropy}}
\end{split}\label{eq:actor}
\end{align}

@@

\begin{figure}[h]
\centering
  \includegraphics[width=0.8\columnwidth]{img/rnn-sampling.pdf}
  \caption{SparseIDS framework. Three deep networks are shown: the classifier, the actor network and the critic network. The actor takes actions based on the feedback from the the critic network and the classifier. The critic network tracks the actor's actions based on the classifier confidence. The packet selector chooses packets based on the decision of the actor network.}
  \label{fig:sampling}
\end{figure}

@@

The architecture in \autoref{fig:neuralNetworkArchitecture} shows that the actor network has two outputs: $\mu$ and $\sigma$. These encode the mean and standard deviation of a log-normal distribution and can never be smaller or equal to zero and hence we always apply a rectifier function at the output. We choose the softplus, which is a smooth rectifier function:
\begin{align}
\text{softplus}(x) = \log\left(1 + e^x \right)
\end{align}
When sampling from the log-normal distribution, a real number is obtained. We round it down and then add 1. This gives us a natural number between 1 and $\infty$, which states how many packets should be skipped.

When experimenting with continuous actions, we encountered one practical problem: At the beginning of training, the classifier would still perform badly and would not give the actor a useful reward. As a result, the actor would only focus on sparsity and aim to maximize it, as the classification reward would always be equally bad no matter how many packets are chosen. Thus, the actor would learn to let $\mu$ go to infinity. This would increase the reward at the beginning when the interaction between the three components of our \gls{rl} system is still fragile. As a solution, we add a penalty to the reward calculation when the actor jumps far behind the last flow packet. The modified reward applies only to the last chosen packet and penalizes the actor jumping more than one packet behind the last packet. The sparsity reward formula (\autoref{eq:R_sparsity}) of the last packet is modified as follows, where \textit{last} refers to the index of the last chosen packet in the flow ($n=\textit{last}$):
\begin{align}
R_{\text{sparsity},\textit{last}} = \frac{\left(\sum_{i=\textit{last}+1}^{N} 1_{\text{skipped},i}\right)}{N- (\textit{last}+1+\left(\textit{last}+a_{\textit{last}} - N - 1\right))}
\label{eq:R_sparsity_last}
\end{align}
Specifically, it adds the term $\left(\textit{last}+a_{\textit{last}} - N - 1\right)$ to the denominator which penalizes the actor jumping more than one packet behind the flow. Jumping just right after the flow does not get penalized because if the actor considers the rest of the flow as irrelevant, ignoring the rest of the flow and jumping to the first packet behind the flow is a useful action.

@@

%\begin{table}[h]
%\caption{Occurrence frequency of the attack flows.}
%\label{tab:occurrence}
%\centering
%\subfloat[CIC-IDS-2017\label{fig:cicids17proportions}
%]{
%\begin{tabular}{l r}
%\toprule
%Attack type & \hspace*{-4mm}Proportion \\ \midrule
%DoS Hulk	&	10.10\%	\\
%PortScan, Firewall	&	6.90\%	\\
%DDoS LOIT	&	4.08\%	\\
%Infiltration	&	3.30\%	\\
%DoS GoldenEye	&	0.32\%	\\
%DoS SlowHTTPTest	&	0.18\%	\\
%DoS Slowloris	&	0.17\%	\\
%Brute-force SSH	&	0.11\%	\\
%Botnet ARES	&	0.03\%	\\
%XSS attack	&	0.03\%	\\
%PortScan, no Fw.	&	0.02\%	\\
%Brute-force FTP	&	0.01\%	\\
%SQL injection	&	$<$0.01\%	\\
%Heartbleed	&	$<$0.01\%	\\
%\bottomrule
%\end{tabular}
%}{}
%\subfloat[UNSW-NB15\label{fig:unswnb15proportions}
%]{
%\begin{tabular}{l r}
%\toprule
%Attack type & \hspace*{-4mm}Proportion \\ \midrule
%Exploits	&	1.42\%	\\
%Fuzzers	&	1.01\%	\\
%Reconnaissance	&	0.58\%	\\
%Generic	&	0.21\%	\\
%DoS	&	0.19\%	\\
%Shellcode	&	0.08\%	\\
%Analysis	&	0.03\%	\\
%Backdoors	&	0.02\%	\\
%Worms	&	0.01\%	\\
%\bottomrule
%\end{tabular}
%}{}
%\end{table}
For our experiments, we use the \textit{CIC-IDS-2017} \cite{sharafaldin_toward_2018} dataset, which includes more than 2 million flows of network data, containing both benign traffic and a large number of different attacks. The distribution of different attacks in the dataset are shown in \autoref{tab:occurrence}.

@@

%\begin{table}
%\caption{Comparing classification metrics per flow with different tradeoffs for both CIC-IDS-2017 and UNSW-NB15.} \label{tab:results_tradeoff}
%\centering
%\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
%& \multicolumn{3}{c}{CIC-IDS-2017} & \multicolumn{3}{c}{UNSW-NB15} \\
%Tradeoff & 0.0 & 0.1 & 0.5 & 0.0 & 0.1 & 0.5 \\	\midrule
%\setrow{\bfseries}Sparsity & 0\% & 76.3\% & 76.5\% & 0\% & 83.5\% & 84.5\% \\ \midrule
%\setrow{\bfseries}Accuracy & 99.5\% & 99.4\% & 99.3\% & 98.9\% & 98.4\% & 98.5\% \\
%Precision & 99.4\% & 98.5\% & 98.7\% & 81.1\% & 69.5\% & 70.8\% \\
%Recall & 98.5\% & 99.0\% & 98.6\% & 84.0\% & 89.9\% & 89.6\% \\
%F1 & 99.0\% & 98.7\% & 98.6\% & 82.5\% & 78.4\% & 79.1\% \\
%Youden & 98.4\% & 98.5\% & 98.2\% & 83.3\% & 88.6\% & 88.4\% \\
%\bottomrule
%\end{tabular*}
%\end{table}
\subsection{Influence of the tradeoff}
\begin{table}[h]
%% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
%\vspace{0.075in}
%\parbox{.45\linewidth}{
\caption{Comparing classification metrics per flow with different tradeoffs. The higher the tradeoff, the higher the sparsity but also the lower the accuracy.} \label{tab:results_tradeoff}
\centering
\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
Tradeoff & 0.0 & 0.1 & 0.5 & 1.0 \\	\midrule
Sparsity & 0\% & 76.3\% & 76.5\% & \textbf{76.8\%} \\ \midrule
Accuracy & \textbf{99.5\%} & 99.4\% & 99.3\% & 99.0\% \\
Precision & \textbf{99.4\%} & 98.5\% & 98.7\% & 97.9\% \\
Recall & 98.5\% & \textbf{99.0\%} & 98.6\% & 98.1\% \\
F1 & \textbf{99.0\%} & 98.7\% & 98.6\% & 98.0\% \\
Youden & 98.4\% & \textbf{98.5\%} & 98.2\% & 97.4\% \\
\bottomrule
\end{tabular*}
%}
\end{table}
\autoref{tab:results_tradeoff} shows the influence of the tradeoff parameter $\alpha$. For columns with a tradeoff of 0.0, we train the classifier regularly with all packets without using \gls{rl}. For tradeoffs 0.1, 0.5 and 1.0 we train \ours{} using \gls{rl} with continuous actions and keep the tradeoff constant for the whole training. Results show that the higher the tradeoff, the fewer packets are sampled and the lower the accuracy.

@@

\begin{table}[h]
%% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
%\vspace{0.075in}
\caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions. The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
\centering
\begin{tabular}{>{\rowmac}l >{\rowmac}r>{\rowmac}r<{\clearrow}} \toprule
Actions & continuous & discrete \\	\midrule
Sparsity & \textbf{76.3\%} & 73.6\% \\ \midrule
Accuracy & \textbf{99.4\%} & 99.2\% \\
Precision & 98.5\% & \textbf{98.7\%} \\
Recall & \textbf{99.0\%} & 98.3\% \\
F1 & \textbf{99.7\%} & 98.5\% \\
Youden & \textbf{98.5\%} & 97.8\% \\
\bottomrule
\end{tabular}
\end{table}
%\begin{table}
%\caption{Comparing classification metrics per flow with either discrete (20 different actions) or continuous actions for CIC-IDS-2017 and UNSW-NB15. The tradeoff is always $0.1$.} \label{tab:results_disc_cont}
%\centering
%\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
%& \multicolumn{2}{c}{CIC-IDS-2017} & \multicolumn{2}{c}{UNSW-NB15} \\
%Actions & cont. & disc. & cont. & disc. \\	\midrule
%\setrow{\bfseries}Sparsity & 76.3\% & 73.6\% & 83.5\% & 86.6\% \\ \midrule
%\setrow{\bfseries}Accuracy & 99.4\% & 99.2\% & 98.4\% & 98.4\% \\
%Precision & 98.5\% & 98.7\% & 69.5\% & 68.0\% \\
%Recall & 99.0\% & 98.3\% & 89.9\% & 94.6\% \\
%F1 & 99.7\% & 98.5\% & 78.4\% & 79.2\% \\
%Youden & 98.5\% & 97.8\% & 88.6\% & 93.2\% \\
%\bottomrule
%\end{tabular*}
%\end{table}
\autoref{tab:results_disc_cont} compares continuous and discrete (20 different) actions and shows that continuous actions result in a higher accuracy with fewer samples being selected. We chose 20 discrete actions as flows have a maximum length of 20, so with 20 actions, the \gls{rl} can decide to skip the entire flow at the first packet. When we tried with fewer discrete actions (like 8), we saw that sparsity was limited as the \gls{rl} would have liked to skip more packets but couldn't. The results lead to the conclusion that at least for our use case and for the dataset we consider, continuous actions seem to perform better. Generally, the advantage of discrete actions is that they are theoretically more powerful as they can compare each of the possible actions with each other, while the only possibility for continuous actions is to increase or decrease the mean of the log-normal distribution, which can lead to worse performance. On the other hand, with discrete actions, the maximum number of samples to skip is fixed, which we consider to be a major drawback. Furthermore, training time is potentially higher since, for example, for a discrete action space of 20 actions, each option is treated as being completely different and, while actually there is not a large difference between skipping 18 or 19 packets, there is a big difference between skipping 0 packets compared to skipping 19 packets. This means that for discrete actions, all actions are perceived as being completely independent, which is not true as the reward function considers sparsity, which increases gradually as more packets are skipped and thus similar actions result in similar rewards.

@@

\begin{table}[h]
% XXX: Max: Without the vspace, the table violates the margins. WTF!!!
\vspace{0.075in}
\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff 0.1) and alternative strategies.}
\label{tab:results_others}
\centering
\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
& \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
Accuracy & \textbf{99.4\%} & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
Precision & \textbf{98.5\%} & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
Recall & \textbf{99.0\%} & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
F1 & \textbf{99.7\%} & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
Youden & \textbf{98.5\%} & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
\bottomrule
\end{tabular*}
%}
\end{table}
%\begin{table}
%\caption{Comparing classification metrics per flow between \gls{rl}-based sampling with continuous actions (tradeoff $0.1$) and alternative strategies.} \label{tab:results_others}
%\centering
%\begin{tabular}{>{\rowmac}l >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
%& \gls{rl} & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
%\setrow{\bfseries}Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
%\setrow{\bfseries}Accuracy & 99.4\% & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
%Precision & 98.5\% & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
%Recall & 99.0\% & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
%F1 & 99.7\% & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
%Youden & 98.5\% & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
%\bottomrule
%\end{tabular}
%\end{table}

Next we compared the \gls{rl} technique used by \ours{} with several common sampling strategies suggested in \autoref{sec:other_sampling_techniques}.
\autoref{tab:results_others} reveals that \ours{} achieves the highest accuracy with the same sparsity. Not surprisingly, random sampling performs the worst, as it is likely that a flow is ``unfortunate'' and doesn't get enough packets sampled while another one gets more packets or the packets that carry more useful information. The best strategy besides the \gls{rl} approach seems to be ``first $m$'' sampling, which matches the conclusion from \cite{hartl_explainability_2020} that the first couple of packets are the most important ones for determining maliciousness of a flow.
``first $m$'' performs better than ``relative first $m$'' because for ``relative first $m$'', longer flows will have more packets sampled from them. Conversely, very short flows will get few packets. For example, with $p=0.2$, a flow of length 20 will get 4 packets while one with 4 packets will get only 1. However, with ``first $m$'', the short and the long flow will get the same number of packets. Apparently it is better if all flows get the same number of packets than if long flows get more packets.
